Towards atomistic resolution structure of phosphatidylcholine glycerol backbone and choline headgroup at different ambient conditions

### 2015-02-17 HGmodelMANUSCRIPT/HGmodel_draft3.tex
update HGmodel manuscript

\subsection{Order parameters from experiments}
\todo{DONE
%We should mention the sign issue and refer to the blog post:
%http://nmrlipids.blogspot.fi/2014/04/on-signs-of-order-parameters.html somewhere.
%Maybe we should change the name of this section II.B. to: "Analysis and terminology of order parameters from experiments 
%and simulations" or similar, and also move the "forking" explanation here.
}
The order parameter of a hydrocarbon C--H vector is defined as 
\begin{equation}\label{orderP}
S_{{\rm CH}}=\frac{3}{2}\langle \cos^2 \theta-1 \rangle,
\end{equation} 
where the angle brackets denote an ensemble average over the sampled conformations, and $\theta$ is the angle between the C--H bond and the membrane normal.
The absolute values of order parameters can be measured by detecting quadrupolar splitting with $^2$H NMR~\cite{seelig77c} or by detecting dipolar 
splitting with $^{13}$C NMR~\cite{hong95a,gross97,dvinskikh05a,ferreira13}. The measurements are based on
different physical interactions and also the connection between order parameters and quadrupolar or dipolar splitting
are different. The absolute values of order parameters from the measured quadrupolar splitting $\Delta \nu_Q$ ($^2$H NMR) are calculated using 
the equation $|S_{{\rm CD}}|=\frac{4}{3} \frac{e^2qQ}{h} \Delta \nu_Q$, where the value for the static quadrupole
splitting constant is estimated from various experiments to be 170~kHz leading to a numerical relation $|S_{{\rm CD}}|=0.00784 \times \Delta \nu_Q$~\cite{seelig77c}. 
The absolute values of order parameters from the measured dipolar splitting $d_\mathrm{CH}$ ($^{13}$C NMR) are calculated using equation
$|S_{{\rm CH}}|=\frac{4\pi\langle r_\mathrm{CH}^3 \rangle}{\hbar \mu_0 \gamma_h \gamma_c} d_\mathrm{CH}$, where
values between 20.2--22.7 kHz are used for $\frac{4\pi\langle r_\mathrm{CH}^3 \rangle}{\hbar \mu_0 \gamma_h \gamma_c}$,
depending on the original authors~\cite{hong95a,gross97,dvinskikh05a,ferreira13}.
It is important to note that the order parameters measured with different techniques based on different physical interactions are in good agreement
with each other (see Results and Discussion), indicating very high quantitative accuracy of the measurements.
For a more detailed discussion see \url{http://nmrlipids.blogspot.fi/2014/02/accuracy-of-order-parameter-measurements.html}
\todo{This is temporarily linked to the blog. Currently I think that the posts (with comments included) to which we want to cite are put into 
the Figshare in pdf format as they are in when the publication is submitted. Figshare allows commenting to continue the discussion.
COMMENT: There is also the option to put the most relevant blog posts as a supplementary material. The advantage of this would
be that the posts would be peer reviewed as well.}.

@@

Slipid \cite{jambeck12}          & POPC & 128 & 5120 & 303 & 200 & 150 & \cite{slipidsFILESpopc}$^*$ & SI \\

@@

\subsection{Full hydration: Experimental order parameters for the glycerol backbone and headgroup}\label{experiments}
The specific deuteration of $\alpha$-, $\beta$- and g$_3$- segments of DPPC has been successful, 
allowing the absolute value order parameter measurements for these segments by $^2$H--NMR~\cite{gally75,brown77,brown78,akutsu81}.
In addition, the absolute values of order parameters for all glycerol backbone and choline headgroup segments in egg yolk lecithin~\cite{hong95a},
DMPC~\cite{hong95b,gross97,dvinskikh05a}, DOPC~\cite{warschawski05} and POPC~\cite{warschawski05,ferreira13}
have been measured with several different implementations of $^{13}$C NMR experiments. In addition, the signs of order parameters in some systems
are measured with $^{13}$C NMR tehcniques~\cite{hong95a,hong95b,gross97}. The experimental absolute values of glycerol backbone 
and choline order parameters from various publications are shown in Fig.~\ref{HGorderparameters}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.6cm]{../DATAreportediINblog/comparisonSorted.pdf}
\todo{Markus Miettinen suggested that we should consider making one more figure where only experimental data would be shown and that would be discussed in Section~\ref{experiments}. \\
COMMENT: I do not think that we need such figure anymore since this figure is more clear now.} \\
\todo{I think that this figure is now very good. There is also the interactive version by Hubert Santuz now in https://plot.ly/~HubertSantuz/72/lipid-force-field-comparison/
we should figure out which is the most practical way to put that behind permalink once it finalized (Zenodo, figshare or something else?) and then put a citation in the paper.}
  \caption{\label{HGorderparameters}
  Order parameteres from simulations listed in Table~\ref{systems} and experiments for glycerol and choline groups.
  The experimental values were taken from the following publications: DMPC 303~K from \cite{gross97}, DMPC 314~K from \cite{dvinskikh05a}, DPPC 310K from \cite{gally75}, 
  DPPC 323~K from \cite{akutsu81}, and POPC 298~K \cite{ferreira13}.
  The vertical bars shown for some of the computational values are not error bars, but demonstrate that for 
  these systems we had two data sets; the ends of the bars mark the values from the two sets, and the dot marks their measurement-time-weighted average. 
  The H{\"o}gberg et al. force field in Table~\ref{systems} is abbreviated in this figure as HoNiLy, Tj{\"o}rnhammar et al. as TjoEdh, CHARMM36-UA as C36UA,
  Rabinovic et al. as RabLyu and Ulmschneider and Ulmschneider as UlmUlm.
} 
\end{figure}

@@

In addition to the numerical values, an important feature of the glycerol backbone is the 
inequality of order parameters for the two hydrogens attached to the same carbon in g$_1$ and g$_3$ segments,
while the hydrogens in choline $\alpha$ and $\beta$ segments give equal values.
The forking in glycerol backbone g$_3$ segment is small ($\approx$ 0.02) 
and some experiments only report the larger value or the average value~\cite{akutsu81,ferreira13}. 
In contrast, forking is significant for the glycerol backbone g$_1$ segment, whose lower order parameter is close to zero and the
larger one has an absolute value of approximately 0.13--0.15. Forking was studied in detail by Gally et al.~\cite{gally81}, who used E. Coli to 
stereospecifically deuterate the different hydrogens attached to the g$_1$ or g$_3$ groups in PE lipids, and measured the order parameters from the lipid 
extract. This experiment gave the lower order parameter when deuterium was in the S position of g$_1$ or R position for g$_3$.
Since the glycerol backbone order parameters are very similar irrespective of the headgroup chemistry (PC,PE and PG) or lipid 
environment~\cite{gally81}, it is reasonable to assume that the stereospecifity measured for the PE lipids
holds also for the PC lipids.

By combining the experimental information of the sign~\cite{hong95a,hong95b,gross97} and the stereospecific
measurements~\cite{gally81} with the accurate absolute value measurements from various techniques~\cite{gally75,akutsu81,gross97,dvinskikh05a,ferreira13},
the most detailed experimentally available order parameter information for the glycerol backbone and choline segments of POPC is obtained.
These data are shown in Fig.~\ref{HGorderparameters2}.
\begin{figure*}[]
  \centering
  \includegraphics[width=8.6cm]{HGorderparameters5.eps} \\
  \todo{Markus Miettinen suggested that we should consider ``spreading'' the values similarly to the figure~\ref{HGorderparameters}} \\
  \caption{\label{HGorderparameters2}
  Order parameteres from simulations with Berger-POPC-07, CHARMM36, GAFFlipid and MacRog force fields together with experimental values for POPC glycerol and choline groups.
  The magnitudes for experimental order parameters are taken from Ferreira et al.~\cite{ferreira13}, the signs are based on the measurements by Hong et al.~\cite{hong95a,hong95b} 
  and Gross et al.~\cite{gross97}, and the R/S labeling is based in the measurements by Gally et al.~\cite{gally81}.
} 
\end{figure*}

@@

To get a general idea of the quality of the glycerol backbone and choline headgroup structures in different models, we calculated the absolute
values of the order parameters for these parts from fourteen different lipid models (Table~\ref{systems}) and 
plotted the results together with experimental values in Fig.~\ref{HGorderparameters}.
Two criteria were used to judge the quality of the model: there must not be significant  {\bf forking} in the $\alpha$ and $\beta$ carbons,
there must be only moderate forking in the g$_3$ carbon and there must be significant forking in the g$_1$ carbon, {\bf absolute values}
should be preferably inside to the subjective sweet spots determined from experiments (blue shaded regions in Fig.~\ref{HGorderparameters}).
None of the studied force fields fulfils these criteria. However, three force fields perform better than others, namely CHARMM36~\cite{klauda10}, MacRog~\cite{maciejewski14} and GAFFlipid~\cite{dickson12}.
The results for each force field in respect to the above criteria are summarized in Figure~\ref{FullHydrationComparisonTable}.
\todo{The table is now done, see Fig.~\ref{FullHydrationComparisonTable}. The text should be updated accordingly.}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.6cm]{../DATAreportediINblog/comparisonTable.pdf}
  \caption{\label{FullHydrationComparisonTable}
  Ranking of force fields based on data of Fig.~\ref{HGorderparameters}.
  } 
\end{figure}

The top three models (CHARMM36, MacRog and GAFFlipid) together with the most used lipid model (Berger based model) 
were subjected to a more careful comparison including the signs and the sterospecific labeling in Fig.~\ref{HGorderparameters2}.
The essential additional information given by this comparison is that the sign of the $\beta$ carbon order parameter is correct only in CHARMM36 model.

@@

The results in the previous section revealed significant differences of the glycerol backbone and choline headgroup
order parameters between different molecular dynamics simulation models.
However, it is not straightforward to conclude which kind of structural differences (if any)
between the models the results indicate, because the mapping from the order parameters to the 
structure is not unique. In this section we demonstrate that 1) the differences in order parameters
indicate significantly different structural sampling strongly correlating with the dihedral angles of the related bonds,
and that 2) the comparison between experimental and simulated order parameters can be used to exclude
nonrealistic structural samping in molecular dynamics simulations. The demonstration is done for 
the dihedral angles defined by the g$_3$-g$_2$-g$_1$-O(\textit{sn}-1) segments in the glycerol backbone and 
the N-$\beta$-$\alpha$-O segments in the headgroup. These dihedrals were chosen for demonstration, because 
significant differences between the models are observed around these segments in Fig.~\ref{HGorderparameters2}.
We note that performing a similar comparison through all the dihedrals in all the 14 models would probably give highly useful
information on how to improve the accuracy of the models yet this is beyond the scope of the current report.

The dihedral angle distributions for the  g$_3$-g$_2$-g$_1$-O(\textit{sn}-1) dihedral calculated from different models are
shown in Fig.~\ref{dihDISTS}. The distribution is qualitatively different for the Berger-POPC-07 model, showing a maximum in 
the gauche$^+$-conformation (60$^o$) compared to all the other models showing a maximum in the anti-conformation (180$^o$).
The distributions in all the other models have the same general features, the main difference being that the
fraction of configurations in the gauche$^-$-conformation (-60$^o$) is zero for the MacRog, detectable for the CHARMM36 and
equally large to the gauche$^+$ fraction in GAFFlipid. From the results we conclude that most likely the wrongly sampled
dihedral angle for the g$_2$-g$_1$ bond explains the significant discrepancy to the experimental order parameters
for the g$_1$ segment in the Berger-POPC-07 model (Fig.~\ref{HGorderparameters2}). 
In conclusion, models preferring the anti conformation for this dihedral give more realistic order parameters and
this in agreement with previous crystal structure and $^1$H NMR studies~\cite{hauser80,hauser81,hauser81b,hauser88,pascher92,marsh06}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.6cm]{g1-g2_Cdihs2.eps}
  \caption{\label{dihDISTS}
    Dihedral angle distributions for g$_3$-g$_2$-g$_1$-O(\textit{sn}-1) dihedral from different models (POPC bilayer in full hydration).
      } 
\end{figure}

@@

\subsection{Response to dehydration and cholesterol content}
In addition to pure phosphatidylcholine bilayers at full hydration, the choline headgroup order parameters
have been measured under various different conditions~\cite{gally75,brown77,brown78,akutsu81,altenbach84,scherer89,bechinger91,ulrich94,dvinskikh05b,castro08,kapla12,ferreira13}.
Also the order parameters for the glycerol backbone have been measured with $^{13}$C NMR in dehydrated conditions~\cite{dvinskikh05b}, and as a function 
of anesthetics~\cite{castro08} and glycolipids~\cite{kapla12} for DMPC, and as a function of cholesterol 
concentration for POPC~\cite{ferreira13}. Due to the high resolution in the NMR (especially $^2$H NMR) experiments,
even very small order parameter changes resulting from the varying conditions can be measured (see
\url{http://nmrlipids.blogspot.fi/2014/02/accuracy-of-order-parameter-measurements.html}
for more discussion.
\todo{This is temporarily linked to the blog. Currently I think that the posts (with comments included) to which we want to cite are put into 
the Figshare in pdf format as they are in when the publication is submitted. Figshare allows commenting to continue the discussion.}
However, as already discussed above, it is not simple to deduce 
the structural changes from order parameter changes~\cite{akutsu91,semchyschyn04}. Consequently, comparison of the order parameters
between simulations and experiments in different conditions can be used to assess the quality of the force field 
in different situations, and, if the quality is good, to potentially interpret the structural changes in experiments.
Here we exemplify such comparison for a lipid bilayer under low hydration levels and when varying amounts of cholesterol is included in the bilayer. 
The interaction between ions and a phosphatidylcholine bilayer will be discussed in a separate study~\cite{ionpaper}.

@@

- None of the tested models (14 different models) produces the order parameters with the experimental
accuracy for a fully hydrated phoshatidylcholine lipid bilayer. However, the CHARMM36, GAFFlipid and MacRog 
models are relatively close. The structures of these models together with the most used lipid model (Berger based) 
were subjected to more careful studies. The results revealed that the current models are not accurate
enough to resolve the atomistic resolution structures sampled by glycerol backbone and choline headgroup.  
However, the correlation between dihedral angle distributions and order parameter differences was found, 
suggesting that careful adjustment of dihedral potentials would potentially lead to the model with a correct
structure.

@@

@@

@@

Initial coordinates for a hydrated DPPC (at 323~K) and POPC (at 310~K) bilayers (30 and 40 waters/lipid, respectively) were taken directly from the Slipids home page \url{http://people.su.se/\~jjm/Stockholm\_Lipids/Downloads.html}.  The Slipids force field~\cite{jambeck12,XXX} ADD 2ND SLIPIDS PAPER! was used for the the all atom descriptions of DPPC and POPC, and
water was described with the TIP3P water model~\cite{jorgensen83}. Simulations were performed within the NPT ensemble using the GROMACS 4.6.X simulation
package~\cite{hess08}. The Nos\'{e}--Hoover thermostat~\cite{nose84,hoover85} was used with reference temperatures of
323~K (DPPC) and 310~K (POPC) and a relaxation time constant of 0.5~ps. Water and lipids were coupled separately to
the heat bath. Pressure was kept constant at 1.013~bar using a semi--isotropic Parrinello--Rahman
barostat~\cite{parrinello81} with a time constant of 10.0~ps. Equations of motion were
integrated with the leapfrog algorithm using a timestep of 2~fs. Long range
electrostatic interactions were calculated using the PME method~\cite{darden93,essman95}, with a fourth order
smoothing spline. A real space cut-off of 1.0~nm was employed with grid spacing of 0.12~nm in the reciprocal space.
Lennard-Jones potentials were cut off at 1.4~nm, with a dispersion correction applied to both energy and pressure. All covalent bonds in lipids were constrained using the LINCS algorithm~\cite{hess97}, 
whereas water molecules were constrained using SETTLE~\cite{miyamoto92}. Twin-range cutoffs,
1.0~nm and 1.6~nm, were used for the neighbor lists with the longrange neighbor list updated every
10 steps. This simulation protocol corresponds to the protocol used in Ref~\cite{jambeck13}. 
%The simulations were run for XXX ns with the last XXX ns used for calculating the order parameters.

>>>>

\subsection{Order parameters from experiments}\label{expORDp}
\todo{DONE
%We should mention the sign issue and refer to the blog post:
%http://nmrlipids.blogspot.fi/2014/04/on-signs-of-order-parameters.html somewhere.
%Maybe we should change the name of this section II.B. to: "Analysis and terminology of order parameters from experiments 
%and simulations" or similar, and also move the "forking" explanation here.
}
The order parameter of a hydrocarbon C--H vector is defined as 
\begin{equation}\label{orderP}
S_{{\rm CH}}=\frac{3}{2}\langle \cos^2 \theta-1 \rangle,
\end{equation} 
where the angle brackets denote an ensemble average over the sampled conformations, and $\theta$ is the angle between the C--H bond and the membrane normal.
The absolute values of order parameters can be measured by detecting quadrupolar splitting with $^2$H NMR~\cite{seelig77c} or by detecting dipolar 
splitting with $^{13}$C NMR~\cite{hong95a,gross97,dvinskikh05a,ferreira13}. The measurements are based on
different physical interactions and also the connection between order parameters and quadrupolar or dipolar splitting
are different. The absolute values of order parameters from the measured quadrupolar splitting $\Delta \nu_Q$ ($^2$H NMR) are calculated using 
the equation $|S_{{\rm CD}}|=\frac{4}{3} \frac{e^2qQ}{h} \Delta \nu_Q$, where the value for the static quadrupole
splitting constant is estimated from various experiments to be 170~kHz leading to a numerical relation $|S_{{\rm CD}}|=0.00784 \times \Delta \nu_Q$~\cite{seelig77c}. 
The absolute values of order parameters from the measured dipolar splitting $d_\mathrm{CH}$ ($^{13}$C NMR) are calculated using equation
$|S_{{\rm CH}}|=\frac{4\pi\langle r_\mathrm{CH}^3 \rangle}{\hbar \mu_0 \gamma_h \gamma_c} d_\mathrm{CH}$, where
values between 20.2--22.7 kHz are used for $\frac{4\pi\langle r_\mathrm{CH}^3 \rangle}{\hbar \mu_0 \gamma_h \gamma_c}$,
depending on the original authors~\cite{hong95a,gross97,dvinskikh05a,ferreira13}.
It is important to note that the order parameters measured with different techniques based on different physical interactions are in good agreement
with each other (see Results and Discussion), indicating very high quantitative accuracy of the measurements.
For a more detailed discussion see \url{http://nmrlipids.blogspot.fi/2014/02/accuracy-of-order-parameter-measurements.html}
\todo{This is temporarily linked to the blog. Currently I think that the posts (with comments included) to which we want to cite are put into 
the Figshare in pdf format as they are in when the publication is submitted. Figshare allows commenting to continue the discussion.
COMMENT: There is also the option to put the most relevant blog posts as a supplementary material. The advantage of this would
be that the posts would be peer reviewed as well.}.

@@

Slipid \cite{jambeck12b}          & POPC & 128 & 5120 & 303 & 200 & 150 & \cite{slipidsFILESpopc}$^*$ & SI \\

@@

\subsection{Full hydration: Experimental order parameters for the glycerol backbone and headgroup}\label{experiments}
{\bf The specific deuteration of $\alpha$-, $\beta$- and g$_3$- segments of DPPC has been successful, 
allowing the absolute value order parameter measurements for these segments by $^2$H--NMR~\cite{gally75,brown77,brown78,akutsu81}.
In addition, the absolute values of order parameters for all glycerol backbone and choline headgroup segments in egg yolk lecithin~\cite{hong95a},
DMPC~\cite{hong95b,gross97,dvinskikh05a}, DOPC~\cite{warschawski05} and POPC~\cite{warschawski05,ferreira13}
have been measured with several different implementations of $^{13}$C NMR experiments. In addition, the signs of order parameters in some systems
are measured with $^{13}$C NMR tehcniques~\cite{hong95a,hong95b,gross97}. The experimental values of glycerol backbone 
and choline order parameters from various publications~\cite{gally75,akutsu81,gross97,dvinskikh05a,ferreira13} with the signs measured in~\cite{hong95a,hong95b,gross97} are shown in Fig.~\ref{HGorderparameters}.
}
\begin{figure}[]
  \centering
  \includegraphics[width=8.6cm]{../DATAreportediINblog/comparisonSorted.pdf}
\todo{Markus Miettinen suggested that we should consider making one more figure where only experimental data would be shown and that would be discussed in Section~\ref{experiments}. \\
COMMENT: I do not think that we need such figure anymore since this figure is more clear now.} \\
\todo{I think that this figure is now very good. There is also the interactive version by Hubert Santuz now in https://plot.ly/~HubertSantuz/72/lipid-force-field-comparison/
we should figure out which is the most practical way to put that behind permalink once it finalized (Zenodo, figshare or something else?) and then put a citation in the paper.}
  \caption{\label{HGorderparameters}
  Order parameteres from simulations listed in Table~\ref{systems} and experiments for glycerol and choline groups.
  The experimental values were taken from the following publications: DMPC 303~K from \cite{gross97}, DMPC 314~K from \cite{dvinskikh05a}, DPPC 310K from \cite{gally75}, 
  DPPC 323~K from \cite{akutsu81}, and POPC 298~K \cite{ferreira13}.
  The vertical bars shown for some of the computational values are not error bars, but demonstrate that for 
  these systems we had two data sets; the ends of the bars mark the values from the two sets, and the dot marks their measurement-time-weighted average. 
  The H{\"o}gberg et al. force field in Table~\ref{systems} is abbreviated in this figure as HoNiLy, Tj{\"o}rnhammar et al. as TjoEdh, CHARMM36-UA as C36UA,
  Rabinovic et al. as RabLyu and Ulmschneider and Ulmschneider as UlmUlm.
} 
\end{figure}

@@

{\bf In addition to the numerical values, an important feature of the glycerol backbone is the 
forking (see section~\ref{expORDp}) of the order parameters in g$_1$ and g$_3$ segments, in contrast to the choline segments $\alpha$ and $\beta$. }
The forking in glycerol backbone g$_3$ segment is small ($\approx$ 0.02) 
and some experiments only report the larger value or the average value~\cite{akutsu81,ferreira13}. 
In contrast, forking is significant for the glycerol backbone g$_1$ segment, whose lower order parameter is close to zero and the
larger one has an absolute value of approximately 0.13--0.15. Forking was studied in detail by Gally et al.~\cite{gally81}, who used E. Coli to 
stereospecifically deuterate the different hydrogens attached to the g$_1$ or g$_3$ groups in PE lipids, and measured the order parameters from the lipid 
extract. This experiment gave the lower order parameter when deuterium was in the S position of g$_1$ or R position for g$_3$.
Since the glycerol backbone order parameters are very similar irrespective of the headgroup chemistry (PC,PE and PG) or lipid 
environment~\cite{gally81}, it is reasonable to assume that the stereospecifity measured for the PE lipids
holds also for the PC lipids.

{\bf By combining the absolute value measurements~\cite{ferreira13}, the sign measurements~\cite{hong95a,hong95b,gross97} and the stereospecific
measurements~\cite{gally81} the most detailed experimentally available order parameter information for the glycerol backbone and choline 
segments of POPC bilayer is collected and the data are shown in Fig.~\ref{HGorderparameters2}. }
\begin{figure*}[]
  \centering
  \includegraphics[width=8.6cm]{HGorderparameters5.eps} \\
  \todo{Markus Miettinen suggested that we should consider ``spreading'' the values similarly to the figure~\ref{HGorderparameters}} \\
  \caption{\label{HGorderparameters2}
  Order parameteres from simulations with Berger-POPC-07, CHARMM36, GAFFlipid and MacRog force fields together with experimental values for POPC glycerol and choline groups.
  The magnitudes for experimental order parameters are taken from Ferreira et al.~\cite{ferreira13}, the signs are based on the measurements by Hong et al.~\cite{hong95a,hong95b} 
  and Gross et al.~\cite{gross97}, and the R/S labeling is based in the measurements by Gally et al.~\cite{gally81}.
} 
\end{figure*}

@@

To get a general idea of the quality of the glycerol backbone and choline headgroup structures in different models, we calculated 
the order parameters for these parts from thirteen different lipid models (Table~\ref{systems}) and 
plotted the results together with experimental values in Fig.~\ref{HGorderparameters}.
Two criteria were used to judge the quality of the model: there must not be significant  {\bf forking} in the $\alpha$ and $\beta$ carbons,
there must be only moderate forking in the g$_3$ carbon and there must be significant forking in the g$_1$ carbon, the {\bf magnitude}
should be preferably inside to the subjective sweet spots determined from experiments (blue shaded regions in Fig.~\ref{HGorderparameters}).
The results for each force field in respect to the above criteria are summarized in Figure~\ref{FullHydrationComparisonTable}.
\todo{DONE}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.6cm]{../DATAreportediINblog/comparisonTable.pdf}
  \caption{\label{FullHydrationComparisonTable}
  Ranking of force fields based on data of Fig.~\ref{HGorderparameters}. 
  Markus, please describe the definition of M, F and $\Sigma$ here.
  } 
\end{figure}

{\bf
None of the studied force fields fulfils these criteria completely, however CHARMM36 is pretty close. 
This is not surprising since the dihedral potentials in this model are tuned to reproduce these parameters better against experiments~\cite{klauda10}.
The next models in the list are CHARMM36-UA~\cite{henin08,lee14} and H\"ogberg08~\cite{hogberg08} which is also not surprising since
these models are using CHARMM bonded potentials for glycerol backbone and choline. The fourth and the fifth models in the list, MacRog~\cite{maciejewski14} and
GAFFlipid~\cite{dickson12}, have independently determined dihedral potentials. All the models based on Gromos potentials and Slipid perform less well.
In the present work we subject the CHARMM36, MacRog, GAFFlipid and Berger-POPC-07 to a more careful comparison including the sterospecific labeling  
(Fig.~\ref{HGorderparameters2}), and atomistic level structure and responses to the dehydration and cholesterol content in the following sections.
These models are selected for more detailed studies since they are the best representatives of different dihedral potential parametrization techniques 
(CHARMM36, MacRog, GAFFlipid), and the Berger based models are the most used lipid model in the literature.
}
%The top three models (CHARMM36, MacRog and GAFFlipid) together with the most used lipid model (Berger based model) 
%were subjected to a more careful comparison including the signs and the sterospecific labeling in Fig.~\ref{HGorderparameters2}.
%The essential additional information given by this comparison is that the sign of the $\beta$ carbon order parameter is correct only in CHARMM36 model.

@@

The results in the previous section revealed significant differences of the glycerol backbone and choline headgroup
order parameters between different molecular dynamics simulation models.
However, it is not straightforward to conclude which kind of structural differences (if any)
between the models the results indicate, because the mapping from the order parameters to the 
structure is not unique. In this section we demonstrate that 1) the differences in order parameters
indicate significantly different structural sampling strongly correlating with the dihedral angles of the related bonds,
and that 2) the comparison between experimental and simulated order parameters can be used to exclude
nonrealistic structural samping in molecular dynamics simulations. The demonstration is done for 
the dihedral angles defined by the g$_3$-g$_2$-g$_1$-O(\textit{sn}-1) segments in the glycerol backbone and 
the N-$\beta$-$\alpha$-O segments in the headgroup. These dihedrals were chosen for demonstration, because 
significant differences between the models are observed around these segments in Fig.~\ref{HGorderparameters2}.
We note that performing a similar comparison through all the dihedrals in all the 13 models would probably give highly useful
information on how to improve the accuracy of the models yet this is beyond the scope of the current report.

The dihedral angle distributions for the  g$_3$-g$_2$-g$_1$-O(\textit{sn}-1) dihedral calculated from different models are
shown in Fig.~\ref{dihDISTS}. The distribution is qualitatively different for the Berger-POPC-07 model, showing a maximum in 
the gauche$^+$-conformation (60$^o$) compared to all the other models showing a maximum in the anti-conformation (180$^o$).
The distributions in all the other models have the same general features, the main difference being that the
fraction of configurations in the gauche$^-$-conformation (-60$^o$) is zero for the MacRog, detectable for the CHARMM36 and
equally large to the gauche$^+$ fraction in GAFFlipid. From the results we conclude that most likely the wrongly sampled
dihedral angle for the g$_2$-g$_1$ bond explains the significant discrepancy to the experimental order parameters
for the g$_1$ segment in the Berger-POPC-07 model (Fig.~\ref{HGorderparameters2}). 
In conclusion, models preferring the anti conformation for this dihedral give more realistic order parameters and
this is in agreement with previous crystal structure and $^1$H NMR studies~\cite{hauser80,hauser81,hauser81b,hauser88,pascher92,marsh06}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.6cm]{g1-g2_Cdihs2.eps}
  \caption{\label{dihDISTS}
    Dihedral angle distributions for g$_3$-g$_2$-g$_1$-O(\textit{sn}-1) dihedral from different models (POPC bilayer in full hydration).
      } 
\end{figure}

@@

\subsection{Response to dehydration and cholesterol content}
In addition to pure phosphatidylcholine bilayers at full hydration, the choline headgroup order parameters
have been measured under various different conditions~\cite{gally75,brown77,brown78,akutsu81,altenbach84,scherer89,bechinger91,ulrich94,dvinskikh05b,castro08,kapla12,ferreira13}.
Also the order parameters for the glycerol backbone have been measured with $^{13}$C NMR in dehydrated conditions~\cite{dvinskikh05b}, and as a function 
of anesthetics~\cite{castro08} and glycolipids~\cite{kapla12} for DMPC, and as a function of cholesterol 
concentration for POPC~\cite{ferreira13}. Due to the high resolution in the NMR (especially $^2$H NMR) experiments,
even very small order parameter changes resulting from the varying conditions can be measured (see
\url{http://nmrlipids.blogspot.fi/2014/02/accuracy-of-order-parameter-measurements.html}
for more discussion.)
\todo{This is temporarily linked to the blog. Currently I think that the posts (with comments included) to which we want to cite are put into 
the Figshare in pdf format as they are in when the publication is submitted. Figshare allows commenting to continue the discussion.}
However, as already discussed above, it is not simple to deduce 
the structural changes from order parameter changes~\cite{akutsu91,semchyschyn04}. Consequently, comparison of the order parameters
between simulations and experiments in different conditions can be used to assess the quality of the force field 
in different situations, and, if the quality is good, to potentially interpret the structural changes in experiments.
Here we exemplify such comparison for a lipid bilayer under low hydration levels and when varying amounts of cholesterol is included in the bilayer. 
The interaction between ions and a phosphatidylcholine bilayer will be discussed in a separate study~\cite{ionpaper}.

@@

- None of the tested models (13 different models) produces the order parameters with the experimental
accuracy for a fully hydrated phoshatidylcholine lipid bilayer. However, the CHARMM36, GAFFlipid and MacRog 
models are relatively close. The structures of these models together with the most used lipid model (Berger based) 
were subjected to more careful studies. The results revealed that the current models are not accurate
enough to resolve the atomistic resolution structures sampled by glycerol backbone and choline headgroup.  
However, the correlation between dihedral angle distributions and order parameter differences was found, 
suggesting that careful adjustment of dihedral potentials would potentially lead to the model with a correct
structure.

@@

{\bf It should be noted that the area per molecule with these settings for the GAFFlipid model was 61.6~\AA$^2$,
while the original publication reported 63.9~\AA$^2$~\cite{dickson12}. However, the same parameters and Amber to Gromacs
conversion reproduced the area per molecule from original publication for the lipid14 model (see next section).}

@@

{\bf The area per molecule with these settings was 65.4~\AA$^2$ which is in agreement with the value reported in the original publciation 65.6$\pm$0.5~\AA$^2$~\cite{dickson14}.}

@@

Initial coordinates for a hydrated DPPC (at 323~K) and POPC (at 310~K) bilayers (30 and 40 waters/lipid, respectively) were taken directly from the 
Slipids home page \url{http://people.su.se/\~jjm/Stockholm\_Lipids/Downloads.html}.  The Slipids force field~\cite{jambeck12,jambeck12b} was used for the the all atom descriptions of DPPC and POPC, and
water was described with the TIP3P water model~\cite{jorgensen83}. Simulations were performed within the NPT ensemble using the GROMACS 4.6.X simulation
package~\cite{hess08}. The Nos\'{e}--Hoover thermostat~\cite{nose84,hoover85} was used with reference temperatures of
323~K (DPPC) and 310~K (POPC) and a relaxation time constant of 0.5~ps. Water and lipids were coupled separately to
the heat bath. Pressure was kept constant at 1.013~bar using a semi--isotropic Parrinello--Rahman
barostat~\cite{parrinello81} with a time constant of 10.0~ps. Equations of motion were
integrated with the leapfrog algorithm using a timestep of 2~fs. Long range
electrostatic interactions were calculated using the PME method~\cite{darden93,essman95}, with a fourth order
smoothing spline. A real space cut-off of 1.0~nm was employed with grid spacing of 0.12~nm in the reciprocal space.
Lennard-Jones potentials were cut off at 1.4~nm, with a dispersion correction applied to both energy and pressure. All covalent bonds in lipids were constrained using the LINCS algorithm~\cite{hess97}, 
whereas water molecules were constrained using SETTLE~\cite{miyamoto92}. Twin-range cutoffs,
1.0~nm and 1.6~nm, were used for the neighbor lists with the longrange neighbor list updated every
10 steps. This simulation protocol corresponds to the protocol used in Ref~\cite{jambeck13}. 
%The simulations were run for XXX ns with the last XXX ns used for calculating the order parameters.
